The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-08-11 paper/iphas-dr2-rev2.tex
Revision 2, submitted 11 Aug 2014

% This is revision 1 of the IPHAS DR2 paper,
% created in response to the referee comments received on 24 June 2014
\documentclass[a4paper,useAMS,usenatbib]{mn2e}
\pdfpagewidth=\paperwidth
\pdfpageheight=\paperheight
\usepackage{amssymb,amsmath}
\usepackage[pdftex]{graphicx}
\usepackage{longtable} 
\usepackage{float}
\usepackage{booktabs}
\usepackage{aas_macros}
\usepackage{caption}
\usepackage{subcaption}

@@

{\bf When the previous release -- the IDR -- was created in late 2007,
just over half of the survey footprint was covered
and the data were insufficiently complete 
to support a homogeneously calibrated source catalogue.
The goal of this paper is to present the next release, 
which supersedes the IDR by including a global re-calibration
and by taking the coverage up to 92 per cent of the survey area.}
In this work we
(i) explain the data reduction 
and quality control procedures that were applied,
(ii) describe and test the new photometric calibration, and 
(iii) detail the construction of the source catalogue
and demonstrate its use.

@@

The detailed properties of the IPHAS observing programme 
have been presented before 
by \citet{Drew2005} and \citet{Gonzalez-Solares2008}. 
To set the stage for this release,
we recap some key points in this section 
{\bf (see Table~1 for a quick-reference overview)}.
IPHAS is an imaging survey of the Galactic Plane north of the celestial equator, 
from which photometry in Sloan $r$ and $i$ 
is extracted along with narrowband \ha. 
It is carried out using the Wide Field Camera (WFC) 
on the 2.5-metre INT in La Palma. 
It is the first digital survey to offer comprehensive optical CCD photometry
of point sources in the Galactic Plane;
the footprint spans a box 
of roughly 180 by 10 degrees, 
covering Galactic latitudes $-5\degr < b < +5\degr$ 
and longitudes $30\degr < \ell < 215\degr$.

@@

The WFC is a mosaic of 4 CCDs 
that captures a sky area of close to 0.29~deg$^2$ at a pixel scale of
0.33 arcsec/pixel.
To cover the Northern Plane with some overlap,
the survey area was divided into 7,635 telescope pointings.
Each of these pointings is accompanied by an offset position
displaced by $+$5 arcmin in Dec 
and $+$5 arcmin in RA,
to deal with inter-CCD gaps, detector imperfections,
and to enable quality checks.
{ \bf An example footprint of a pointing
and its offset position is shown in Fig.~\ref{fig:fieldpair}.}
Hence, the basic unit of observation
amounts to $2 \times 3$ exposures, 
in which each of the 3 survey filters is exposed at 2 offset sky positions 
within, typically, an elapsed time of 10 minutes.
We shall refer to the unit of 3 exposures at the same position 
as a \emph{field},
and the combination of two fields at a small offset as a \emph{field pair}.
Altogether, the survey contains 15,270 fields
grouped into 7,635 field pairs.
To achieve the desired survey depth
of 20th magnitude or fainter in each filter, 
the exposure times were set at 120~s (\ha), 
30~s ($r$) and 10~s ($i$)
in the vast majority of the survey observations.\footnote{The $r$-band exposure time was 10s instead of 30s in the first months of data taking. Since October 2010, the $i$-band exposure time 
has been increased from 10s to 20s to by-pass a sporadic exposure timing bug that affects the WFC.}

@@

All raw IPHAS data were transferred to the Cambridge Astronomical Survey Unit 
(CASU) for initial processing and archiving.  The procedures used by CASU 
were originally devised for the INT Wide Field imaging Survey 
\citep[WFS;][]{McMahon2001,Irwin2005}, which was a 200 deg$^2$ 
{\bf extragalactic} survey programme carried out between 1998 and 2003.
Because IPHAS uses the same telescope and camera combination, we have been 
able to benefit from the existing WFS pipeline. A description of the 
processing steps can be found in \citet{Irwin2001}. Its application to IPHAS 
has previously been described
by \citet{Drew2005} and \citet{Gonzalez-Solares2008},
and some elements of the the source code are available 
on line\footnote{http://casu.ast.cam.ac.uk/surveys-projects/software-release}. 
In brief, the imaging processing part of the pipeline takes care of bias 
subtraction, the linearity correction, flat-fielding with internal gain
correction, and de-fringing for the i-band.

{ \bf
Object detection and parametrisation is then carried out
using the standard methods developed by CASU,
which can be summarised in four steps
\citep[a discussion on each of these steps
and related points can be found in][]{Irwin1985,Irwin1997}:
\begin{enumerate}
\item The local sky background is estimated by first computing an iteratively
sigma-clipped median intensity on a grid of 64$\times$64 pixel bins across 
the image from each detector.  This is usually robust against contaminating 
sources corrupting the background level. The resulting array of background
values is then filtered to further reduce the effect of large objects on
the local background level. Bilinear interpolation is then used to obtain an 
estimate of the background level at the orginal image pixel scale.
\item To improve faint object detection, each image is smoothed with a matched
detection filter which is used in conjunction with the unsmoothed image
for object detection and parameterisation.
\item Objects, or blends of objects, are detected
by identifying groups of 4 or more neighbouring contiguous pixels
in which the intensity exceeds the background level by at least $1.25\sigma$
on the matched filtered image. Objects are deblended using a sequence of
successively higher detection levels.
\item The objects are parametrised using the unsmoothed image at these pixel 
locations : positions are obtained based on an intensity-weighted isophotal 
centre-of-gravity of each object; whilst photometry is derived by measuring 
the intensity in a series of soft-edged circular apertures covering a range 
of diameters (1\farcs 2, 2\farcs 3, 3\farcs 3, 4\farcs 6, 6\farcs 6 and 
9\farcs 2).  Objects are also classified morphologically
-- stellar, extended or noise -- based on their intensity as a function of 
aperture size and on their intensity-weighted second moments, where the latter
are used to derive an estimate of object ellipticity.
\item The parametrisation of overlapping objects is refined by simultaneous 
fitting of soft-edged apertures to each blend,
effectively carrying out ``top hat'' profile fitting where it is necessary.
We note that the parametrisation of these blended objects is naturally
less reliable than that of single, unconfused sources,
which is why they are flagged in the catalogue
(to be explained in \S\ref{sec:catalogue}).
\end{enumerate}

Having carried out object detection,
the astrometric calibration is then determined.
The solution starts with a rough 
World Coordinate System (WCS)
based on the known telescope and camera geometry,
using a Zenithal Polynomial projection
\citep[ZPN;][]{Calabretta2002}
to model the (fixed) field distortion of the camera.
The parameters of this solution
are then progressively refined by fitting against the
Two-Micron All Sky Survey \citep[2MASS; ][]{Skrutskie2006}.
We did not include proper motion information during this step,
because each CCD typically contains
hundreds of 2MASS reference stars.
The resulting fit has previously been shown
to deliver results which are internally consistent
to better than 0\farcs 1 across the detector array
\citep{Gonzalez-Solares2008}.

An external validation of the astrometry 
has been carried out by comparing our positions against the
United States Naval Observatory CCD Astrograph Catalog 
\citep[UCAC4;][]{UCAC4}.
Fig.~\ref{fig:astrometry} shows the distribution
of the offsets against UCAC4
computed for 1.3 million stars
in the magnitude range $13 < r < 15$,
which is the range where both surveys overlap
and where the formal mean error of UCAC4
is better than 50~mas \citep{UCAC3}.
We find the mean position offset to be $0\farcs 09\pm0\farcs 07$,
which is satisfactory for our purposes.

\begin{figure}
    \includegraphics[width=\linewidth]{figures/astrometry/astrometry.pdf} 
    \caption{Distribution of the astrometric residuals of stars
    		which appear both in IPHAS DR2 and UCAC4
    		within a cross-matching distance of 1\arcsec.
            The residuals were computed for 
            the 1.3 million stars in the IPHAS catalogue
            which are not blended, not saturated,
            and fall in the magnitude range $13 < r < 15$.
            The mean and standard deviation of this distribution 
            equals $0\farcs 09\pm0\farcs 07$.}
    \label{fig:astrometry}
\end{figure}

@@

The entire data set -- comprising 2.5~terabyte of FITS files --
was then transferred to the University of Hertfordshire
for the purpose of transforming the raw
detection tables into a source catalogue which
(i) is quality-controlled,
(ii) is homogeneously calibrated, and 
(iii) contains user-friendly columns and warning flags.
It is these post-processing steps which distinguish this release
from the IDR, which (i) enforced less stringent quality limits,
(ii) did not offer a global calibration,
and (iii) provided a catalogue in which duplicate detections
of unique sources were not flagged.
These improvements are explained next.}

@@

(2) \emph{Ellipticity.} 
The ellipticity of a point source,
defined as $e = 1 - b / a$ 
with $b$ the semi-minor and $a$ the semi-major axis,
is a morphological measure of the elongation
of the Point Spread Function (PSF).
It is expected to be zero (circular)
in a perfect noise-free imaging system,
but it is slightly non-zero in any real telescope data 
due to optical distortions, tracking errors and photon plus readout noise.
Indeed, it is worth noting
that IPHAS data have been collected
from unguided exposures that rely entirely
on the INT's tracking capability.
The mean ellipticity measured
in the data is $0.09\pm0.04$.
There have been sporadic episodes with higher ellipticities
due to mechanical glitches in the telescope tracking system.
{ \bf As a result, 3 per cent of our images show an
average ellipticity across the detectors which is worse
than $e > 0.2$.
The inspection of these examples revealed no evidence
for degraded photometry up to ellipticities of 0.3.
We have excluded a small number of exposures
which exceeded $e = 0.3$.}

@@

For the purpose of providing an initial calibration 
of the $r$ and $i$ broadband fluxes,
photometric standard fields were observed every night.
The standards were chosen from a list based on 
the \cite{Landolt1992} and Stetson (http://cadcwww.dao.nrc.ca/standards) 
objects.
Two or three standard fields were observed 
during the evening and morning twilight,
and at intervals of 2-3 hours throughout the night.
The CASU pipeline automatically identified the observed standards 
and used them to determine a sigma-clipped average zeropoint \textsc{magzpt}
for each night and filter,
such that the number counts $DN$ 
in the pipeline-corrected CCD frames
relate to a magnitude $m$ as:
\begin{equation}
\begin{split}
   m  = & \textsc{magzpt} - 2.5 \log_{10}( DN / \textsc{exptime} ) \\
 &  - \textsc{extinct}\cdot(\textsc{airmass}-1) - \textsc{apcor} - \textsc{percorr},
\label{eqn:mag}
\end{split}
\end{equation}
where \textsc{exptime} is the exposure time in seconds,
\textsc{extinct} is the atmospheric extinction coefficient 
(set in the pipeline at 0.09 for $r$ and 0.05 for $i$ as representative
averages for the telescope site),
\textsc{airmass} is the normalised optical path length 
through the atmosphere and
\textsc{apcor} is a correction for the flux
lost outside of the aperture used
(we adopt a 2\farcs 3-diameter circular aperture by default).
{\bf
Finally, \textsc{percorr} is a term used to correct 
for the small difference in internal gain
computed using the relatively blue twilight flats
compared to the much redder typical astronomical objects. 
It is estimated by making a robust average of the dark sky levels 
measured on each detector during an observing run
(the correction is $0.01\pm0.01$ on average in $i$ and
$0.00\pm0.00$ in $r$ and \ha).
All these quantities correspond to header keywords
in the FITS files produced by the CASU pipeline. }

@@

Despite the best efforts made to obtain a nightly calibration,
large surveys naturally possess field-to-field variations
due to atmospheric changes during the night
and imperfections in the pipeline or the instrument
(e.g. the WFC is known to suffer from sporadic errors
in the timing of exposures).
{ \bf
This is demonstrated in Fig.~\ref{fig:ccd_before},
where we show the combined colour-colour diagram
for nearly 3,000 fields across an area of 400~deg$^2$.
The main locus of stars is poorly defined
in the diagram due to the presence of
incorrectly calibrated fields,
which need to be corrected
during a global re-calibration step.
The application of such a procedure,
to be explained below,
has revealed that the error in our initial nightly calibration
exceeded 0.1~mag in 12 per cent of the fields,
and even exceeded 0.5~mag in 0.7 per cent.
Fig.~\ref{fig:ccd_after}
demonstrates the improvement 
in the colour-colour diagram
after re-calibrating. }

@@

Surveys which observe identical stars at different epochs
can use the repeat measurements to ensure a homogeneous calibration.
{\bf For example, 2MASS attained its global calibration
by observing two standard fields each hour, }
allowing zeropoint variations to be tracked 
over short timescales \citep{Nikolaev2000}.
Alternatively, the SDSS and PanSTARRS surveys could benefit
from revisiting regions in their footprint to 
carry out a so-called \emph{ubercalibration}\footnote{`ubercalibration'
refers to the name of the code used to re-calibrate SDSS photometry. 
It is an anglicised version of the German word `\"uberkalibration',
which was reportedly chosen because the initial authors, Schlegel and Finkbeiner, both
have German-sounding names \citep{Finkbeiner2010}.} procedure,
in which repeat measurements of stars in different nights
are used to fit the calibration parameters
\citep{Ivezic2007,Padmanabhan2008,Schlafly2012}.

@@

Having re-calibrated all fields to the expected APASS accuracy of 3 per cent,
we turned to a different survey, SDSS Data Release 9 \citep{Ahn2012},
to validate the results.
{\bf SDSS DR9 includes photometry across several strips at low
Galactic latitudes,
which were observed as part of 
the Sloan Extension for Galactic Understanding and Exploration \citep[SEGUE;][]{segue}.}
These strips provide data across 18 per cent of the fields in our data release.
We cross-matched the IPHAS fields against the subset of
objects marked as reliable stars in the SDSS catalogue\footnote{
We used the CasJobs facility located at http://skyserver.sdss3.org/CasJobs
to obtain photometry from the SDSS {\sc photoprimary} table 
with criteria {\sc type = star}, {\sc clean = 1} and {\sc score $>$ 0.7}.}
in much the same way as for APASS,
with the difference of selecting from the  fainter magnitude ranges of 
$15<r_{\rm SDSS}<18.0$ and $14.5<i_{\rm SDSS}<17.5$.
This provided us with a set of 1.2~million cross-matched stars.

@@

{\bf
The second step in compiling the source catalogue
is to merge the contemporaneous trios
of $r$, $i$, \ha\ detection tables
into multi-band field catalogues.
This required a position matching procedure 
to link sources between the three bands.
We used the TMATCHN function 
of the STILTS software
for this purpose,
which allows rows from multiple tables
to be matched \citep{Taylor2006}.
In brief, the algorithm
identifies groups of linked detections such that
(i) each detection is located within a specified maximum distance from
one or more members of the group,
(ii) each detection appears in exactly one group, and
(iii) the largest possible groups are preferred
(i.e. preferably containing three detections from all three bands).
The result of the procedure is a band-merged catalogue
in which each row corresponds
to a group of linked $r$, $i$, and \ha\ detections
which satisfy the matching distance criterion
in a pair-wise sense.
Sources for which no counterpart was identified
in all three bands are retained in the catalogue
as single- or double-band detections,
with empty columns for the missing bands.

@@

The resulting band-merged catalogues were inspected by eye
as part of our quality control procedures
and were found to be reliable for the vast majority of objects.
We do warn that blended objects
can occasionally fall victim to source confusion
during the band-merging procedure,
which is a complicated problem
that we have not attempted to resolve in this release.
It is important to bear this in mind when appraising stars
of seemingly unusual colours (such as candidate emission line stars).
If blending is flagged, or if the inter-band matching distance
is unusually large, then the probability that the unusual colour is 
spurious due to source confusion is greatly increased.
}

@@

As the final step, the primary detections
selected above were compiled
into the final 99-column source catalogue
that is described in Appendix~\ref{app:columns}.
{ \bf
The original unweeded list of sources naturally included 
a significant number of spurious entries
as a result of the sensitive detection level
that is employed by the CASU pipeline.
We have decided to enforce three basic criteria
which must be met for a candidate source
to be included in the catalogue:}
\begin{enumerate}
\item the source must have been detected at S/N$>5$ in at least
one of the bands, i.e. it is required that at least one of
\emph{rErr}, \emph{iErr} or \emph{haErr} is smaller
than 0.2 mag;
\item the shape of the source must not be an obvious
cosmic ray or noise artefact, i.e. we require
either \emph{pStar} or \emph{pGalaxy} to be
greater than 20 per cent;
\item the source must not have been detected
in one of the strongly vignetted corners of the instrument, 
not have had any known bad pixels in the aperture,
and not have been on the edge of one of the CCDs.
\end{enumerate}

@@

Table~\ref{tbl:detections} details the breakdown of these
sources as a function of the bands in which they are captured.
159 million sources are detected in all three filters (73 per cent),
30 million are detected in two filters (14 per cent),
and the remaining 30 million are single-band detections.
{\bf
Table~\ref{tbl:detections} also presents 
the fraction of `confirmed' objects,
which we define as those sources
which have been detected more than once
(recall that much of the survey area
is observed twice due to the field pair strategy).
We find that the single-band detections tend
to show much lower confirmation rates (32 per cent on average)
than double- and triple-band detections (89 per cent).
This suggests that a significant fraction of
these entries may be spurious detections.

@@

Not all the single-band detections are spurious, however.
We note that the confirmation rate for $i$-band detections
is markedly better than for $r$ and \ha,
which is likely to be explained by the fact that $i$ is least
affected by interstellar extinction, and so the survey can occasionally pick up
highly-reddened objects in $i$ which are otherwise lost in $r$ and \ha.
Moreover, objects which are intrinsically very red may also be picked
up in $i$ alone, while faint objects with very strong Balmer emission
may appear only in \ha.
Nevertheless, we recommend users not to rely on single-band objects
without inspecting the image data by eye,
or verifying that the object
was detected more than once (\emph{nObs}$>$1).
}

@@

{\bf

@@

{\bf Towards Sh2-82,
the IPHAS and Spitzer/GLIMPSE catalogues
have 4,798 entries in common,
out of 10,739 and 11,321 entries in total, respectively.
Fig.~\ref{fig:spitzer-ccd} presents the Spitzer colour-colour diagram
of the region, showing the 1,356 objects that were detected 
in all four GLIMPSE bands.
We have indicated the 845 objects (62 per cent)
that have a counterpart in IPHAS (grey circles)
and the 511 objects that do not (red crosses).

>>>>

% This is revision 2 of the IPHAS DR2 paper,
% created in response to the referee comments received on 11 August 2014
\documentclass[a4paper,useAMS,usenatbib]{mn2e}
\pdfpagewidth=\paperwidth
\pdfpageheight=\paperheight
\usepackage{amssymb,amsmath}
\usepackage[pdftex]{graphicx}
\usepackage{longtable} 
\usepackage{float}
\usepackage{booktabs}
\usepackage{aas_macros}
\usepackage{caption}
\usepackage{subcaption}

@@

When the previous release -- the IDR -- was created in late 2007,
just over half of the survey footprint was covered
and the data were insufficiently complete 
to support a homogeneously calibrated source catalogue.
The goal of this paper is to present the next release, 
which supersedes the IDR by including a global re-calibration
and by taking the coverage up to 92 per cent of the survey area.
In this work we
(i) explain the data reduction 
and quality control procedures that were applied,
(ii) describe and test the new photometric calibration, and 
(iii) detail the construction of the source catalogue
and demonstrate its use.

@@

The detailed properties of the IPHAS observing programme 
have been presented before 
by \citet{Drew2005} and \citet{Gonzalez-Solares2008}. 
To set the stage for this release,
we recap some key points in this section 
(see Table~1 for a quick-reference overview).
IPHAS is an imaging survey of the Galactic Plane north of the celestial equator, 
from which photometry in Sloan $r$ and $i$ 
is extracted along with narrowband \ha. 
It is carried out using the Wide Field Camera (WFC) 
on the 2.5-metre INT in La Palma. 
It is the first digital survey to offer comprehensive optical CCD photometry
of point sources in the Galactic Plane;
the footprint spans a box 
of roughly 180 by 10 degrees, 
covering Galactic latitudes $-5\degr < b < +5\degr$ 
and longitudes $30\degr < \ell < 215\degr$.

@@

The WFC is a mosaic of 4 CCDs 
that captures a sky area of close to 0.29~deg$^2$ at a pixel scale of
0.33 arcsec/pixel.
To cover the Northern Plane with some overlap,
the survey area was divided into 7,635 telescope pointings.
Each of these pointings is accompanied by an offset position
displaced by $+$5 arcmin in Dec 
and $+$5 arcmin in RA,
to deal with inter-CCD gaps, detector imperfections,
and to enable quality checks.
An example footprint of a pointing
and its offset position is shown in Fig.~\ref{fig:fieldpair}.
Hence, the basic unit of observation
amounts to $2 \times 3$ exposures, 
in which each of the 3 survey filters is exposed at 2 offset sky positions 
within, typically, an elapsed time of 10 minutes.
We shall refer to the unit of 3 exposures at the same position 
as a \emph{field},
and the combination of two fields at a small offset as a \emph{field pair}.
Altogether, the survey contains 15,270 fields
grouped into 7,635 field pairs.
To achieve the desired survey depth
of 20th magnitude or fainter in each filter, 
the exposure times were set at 120~s (\ha), 
30~s ($r$) and 10~s ($i$)
in the vast majority of the survey observations.\footnote{The $r$-band exposure time was 10s instead of 30s in the first months of data taking. Since October 2010, the $i$-band exposure time 
has been increased from 10s to 20s to by-pass a sporadic exposure timing bug that affects the WFC.}

@@

All raw IPHAS data were transferred to the Cambridge Astronomical Survey Unit 
(CASU) for initial processing and archiving.  The procedures used by CASU 
were originally devised for the INT Wide Field imaging Survey 
\citep[WFS;][]{McMahon2001,Irwin2005}, which was a 200 deg$^2$ 
extragalactic survey programme carried out between 1998 and 2003.
Because IPHAS uses the same telescope and camera combination, we have been 
able to benefit from the existing WFS pipeline. A description of the 
processing steps can be found in \citet{Irwin2001}. Its application to IPHAS 
has previously been described
by \citet{Drew2005} and \citet{Gonzalez-Solares2008},
and some elements of the the source code are available 
on line\footnote{http://casu.ast.cam.ac.uk/surveys-projects/software-release}. 
In brief, the imaging processing part of the pipeline takes care of bias 
subtraction, the linearity correction, flat-fielding with internal gain
correction, and de-fringing for the i-band.

Object detection and parametrisation is then carried out
using the standard methods developed by CASU,
which can be summarised in four steps
\citep[a discussion on each of these steps
and related points can be found in][]{Irwin1985,Irwin1997}:
\begin{enumerate}
\item The local sky background is estimated by first computing an iteratively
sigma-clipped median intensity on a grid of 64$\times$64 pixel bins across 
the image from each detector.  This is usually robust against contaminating 
sources corrupting the background level. The resulting array of background
values is then filtered to further reduce the effect of large objects on
the local background level. Bilinear interpolation is then used to obtain an 
estimate of the background level at the orginal image pixel scale.
\item To improve faint object detection, each image is smoothed with a matched
detection filter which is used in conjunction with the unsmoothed image
for object detection and parameterisation.
\item Objects, or blends of objects, are detected
by identifying groups of 4 or more neighbouring contiguous pixels
in which the intensity exceeds the background level by at least $1.25\sigma$
on the matched filtered image. Objects are deblended using a sequence of
successively higher detection levels.
\item The objects are parametrised using the unsmoothed image at these pixel 
locations : positions are obtained based on an intensity-weighted isophotal 
centre-of-gravity of each object; whilst photometry is derived by measuring 
the intensity in a series of soft-edged circular apertures covering a range 
of diameters (1\farcs 2, 2\farcs 3, 3\farcs 3, 4\farcs 6, 6\farcs 6 and 
9\farcs 2).  Objects are also classified morphologically
-- stellar, extended or noise -- based on their intensity as a function of 
aperture size and on their intensity-weighted second moments, where the latter
are used to derive an estimate of object ellipticity.
\item The parametrisation of overlapping objects is refined by simultaneous 
fitting of soft-edged apertures to each blend,
effectively carrying out ``top hat'' profile fitting where it is necessary.
We note that the parametrisation of these blended objects is naturally
less reliable than that of single, unconfused sources,
which is why they are flagged in the catalogue
(to be explained in \S\ref{sec:catalogue}).
\end{enumerate}

Having carried out object detection,
the astrometric calibration is then determined.
The solution starts with a rough 
World Coordinate System (WCS)
based on the known telescope and camera geometry,
using a Zenithal Polynomial projection
\citep[ZPN;][]{Calabretta2002}
to model the (fixed) field distortion of the camera.
The parameters of this solution
are then progressively refined by fitting against the
Two-Micron All Sky Survey \citep[2MASS; ][]{Skrutskie2006},
{ \bf
albeit without correcting for the $\sim$10 years of proper motion
between the IPHAS and 2MASS epochs. }
The resulting fit has previously been shown
to deliver results which are internally consistent
to better than 0\farcs 1 across the detector array
\citep{Gonzalez-Solares2008}.

An external validation of the astrometry 
has been carried out by comparing our positions against the
United States Naval Observatory CCD Astrograph Catalog 
\citep[UCAC4;][]{UCAC4}.
Fig.~\ref{fig:astrometry} shows the distribution
of the astrometric offsets
computed for 1.3 million stars
in the magnitude range $13 < r < 15$,
which is the range where both surveys overlap
and where the formal mean error of UCAC4
is better than 50~mas \citep{UCAC3}.
We find the mean difference in position 
between IPHAS and UCAC4 to be 94~mas,
which is satisfactory for our purposes.
{ \bf
The remaining residuals are in part due to
the motion of the Earth through our galaxy,
which we did not account for.
}

\begin{figure}
    \includegraphics[width=\linewidth]{figures/astrometry/astrometry.pdf} 
    \caption{Distribution of the astrometric residuals of stars
    		which appear both in IPHAS DR2 and UCAC4
    		within a cross-matching distance of 1\arcsec.
            The residuals were computed for 
            the 1.3 million stars in the IPHAS catalogue
            which are not blended, not saturated,
            and fall in the magnitude range $13 < r < 15$.
            The mean and standard deviation of this distribution 
            equals $94\pm65$~mas.}
    \label{fig:astrometry}
\end{figure}

@@

The entire data set -- comprising 2.5~terabyte of FITS files --
was then transferred to the University of Hertfordshire
for the purpose of transforming the raw
detection tables into a source catalogue which
(i) is quality-controlled,
(ii) is homogeneously calibrated, and 
(iii) contains user-friendly columns and warning flags.
It is these post-processing steps which distinguish this release
from the IDR, which (i) enforced less stringent quality limits,
(ii) did not offer a global calibration,
and (iii) provided a catalogue in which duplicate detections
of unique sources were not flagged.
These improvements are explained next.

@@

(2) \emph{Ellipticity.} 
The ellipticity of a point source,
defined as $e = 1 - b / a$ 
with $b$ the semi-minor and $a$ the semi-major axis,
is a morphological measure of the elongation
of the Point Spread Function (PSF).
It is expected to be zero (circular)
in a perfect noise-free imaging system,
but it is slightly non-zero in any real telescope data 
due to optical distortions, tracking errors and photon plus readout noise.
Indeed, it is worth noting
that IPHAS data have been collected
from unguided exposures that rely entirely
on the INT's tracking capability.
The mean ellipticity measured
in the data is $0.09\pm0.04$.
There have been sporadic episodes with higher ellipticities
due to mechanical glitches in the telescope tracking system.
As a result, 3 per cent of our images show an
average ellipticity across the detectors which is worse
than $e > 0.2$.
The inspection of these examples revealed no evidence
for degraded photometry up to ellipticities of 0.3.
We have excluded a small number of exposures
which exceeded $e = 0.3$.

@@

For the purpose of providing an initial calibration 
of the $r$ and $i$ broadband fluxes,
photometric standard fields were observed every night.
The standards were chosen from a list based on 
the \cite{Landolt1992} and Stetson (http://cadcwww.dao.nrc.ca/standards) 
objects.
Two or three standard fields were observed 
during the evening and morning twilight,
and at intervals of 2-3 hours throughout the night.
The CASU pipeline automatically identified the observed standards 
and used them to determine a sigma-clipped average zeropoint \textsc{magzpt}
for each night and filter,
such that the number counts $DN$ 
in the pipeline-corrected CCD frames
relate to a magnitude $m$ as:
\begin{equation}
\begin{split}
   m  = & \textsc{magzpt} - 2.5 \log_{10}( DN / \textsc{exptime} ) \\
 &  - \textsc{extinct}\cdot(\textsc{airmass}-1) - \textsc{apcor} - \textsc{percorr},
\label{eqn:mag}
\end{split}
\end{equation}
where \textsc{exptime} is the exposure time in seconds,
\textsc{extinct} is the atmospheric extinction coefficient 
(set in the pipeline at 0.09 for $r$ and 0.05 for $i$ as representative
averages for the telescope site),
\textsc{airmass} is the normalised optical path length 
through the atmosphere and
\textsc{apcor} is a correction for the flux
lost outside of the aperture used
(we adopt a 2\farcs 3-diameter circular aperture by default).
Finally, \textsc{percorr} is a term used to correct 
for the small difference in internal gain
computed using the relatively blue twilight flats
compared to the much redder typical astronomical objects. 
It is estimated by making a robust average of the dark sky levels 
measured on each detector during an observing run
(the correction is $0.01\pm0.01$ on average in $i$ and
$0.00\pm0.00$ in $r$ and \ha).
All these quantities correspond to header keywords
in the FITS files produced by the CASU pipeline.

@@

Despite the best efforts made to obtain a nightly calibration,
large surveys naturally possess field-to-field variations
due to atmospheric changes during the night
and imperfections in the pipeline or the instrument
(e.g. the WFC is known to suffer from sporadic errors
in the timing of exposures).
This is demonstrated in Fig.~\ref{fig:ccd_before},
where we show the combined colour-colour diagram
for nearly 3,000 fields across an area of 400~deg$^2$.
The main locus of stars is poorly defined
in the diagram due to the presence of
incorrectly calibrated fields,
which need to be corrected
during a global re-calibration step.
The application of such a procedure,
to be explained below,
has revealed that the error in our initial nightly calibration
exceeded 0.1~mag in 12 per cent of the fields,
and even exceeded 0.5~mag in 0.7 per cent.
Fig.~\ref{fig:ccd_after}
demonstrates the improvement 
in the colour-colour diagram
after re-calibrating.

@@

Surveys which observe identical stars at different epochs
can use the repeat measurements to ensure a homogeneous calibration.
For example, 2MASS attained its global calibration
by observing two standard fields each hour,
allowing zeropoint variations to be tracked 
over short timescales \citep{Nikolaev2000}.
Alternatively, the SDSS and PanSTARRS surveys could benefit
from revisiting regions in their footprint to 
carry out a so-called \emph{ubercalibration}\footnote{`ubercalibration'
refers to the name of the code used to re-calibrate SDSS photometry. 
It is an anglicised version of the German word `\"uberkalibration',
which was reportedly chosen because the initial authors, Schlegel and Finkbeiner, both
have German-sounding names \citep{Finkbeiner2010}.} procedure,
in which repeat measurements of stars in different nights
are used to fit the calibration parameters
\citep{Ivezic2007,Padmanabhan2008,Schlafly2012}.

@@

Having re-calibrated all fields to the expected APASS accuracy of 3 per cent,
we turned to a different survey, SDSS Data Release 9 \citep{Ahn2012},
to validate the results.
SDSS DR9 includes photometry across several strips at low
Galactic latitudes,
which were observed as part of 
the Sloan Extension for Galactic Understanding and Exploration \citep[SEGUE;][]{segue}.
These strips provide data across 18 per cent of the fields in our data release.
We cross-matched the IPHAS fields against the subset of
objects marked as reliable stars in the SDSS catalogue\footnote{
We used the CasJobs facility located at http://skyserver.sdss3.org/CasJobs
to obtain photometry from the SDSS {\sc photoprimary} table 
with criteria {\sc type = star}, {\sc clean = 1} and {\sc score $>$ 0.7}.}
in much the same way as for APASS,
with the difference of selecting from the  fainter magnitude ranges of 
$15<r_{\rm SDSS}<18.0$ and $14.5<i_{\rm SDSS}<17.5$.
This provided us with a set of 1.2~million cross-matched stars.

@@

The second step in compiling the source catalogue
is to merge the contemporaneous trios
of $r$, $i$, \ha\ detection tables
into multi-band field catalogues.
This required a position matching procedure 
to link sources between the three bands.
We used the TMATCHN function 
of the STILTS software
for this purpose,
which allows rows from multiple tables
to be matched \citep{Taylor2006}.
In brief, the algorithm
identifies groups of linked detections such that
(i) each detection is located within a specified maximum distance from
one or more members of the group,
(ii) each detection appears in exactly one group, and
(iii) the largest possible groups are preferred
(i.e. preferably containing three detections from all three bands).
The result of the procedure is a band-merged catalogue
in which each row corresponds
to a group of linked $r$, $i$, and \ha\ detections
which satisfy the matching distance criterion
in a pair-wise sense.
Sources for which no counterpart was identified
in all three bands are retained in the catalogue
as single- or double-band detections,
with empty columns for the missing bands.

@@

The resulting band-merged catalogues were inspected by eye
as part of our quality control procedures
and were found to be reliable for the vast majority of objects.
We do warn that blended objects
can occasionally fall victim to source confusion
during the band-merging procedure,
which is a complicated problem
that we have not attempted to resolve in this release.
It is important to bear this in mind when appraising stars
of seemingly unusual colours (such as candidate emission line stars).
If blending is flagged, or if the inter-band matching distance
is unusually large, then the probability that the unusual colour is 
spurious due to source confusion is greatly increased.

@@

As the final step, the primary detections
selected above were compiled
into the final 99-column source catalogue
that is described in Appendix~\ref{app:columns}.
The original unweeded list of sources naturally included 
a significant number of spurious entries
as a result of the sensitive detection level
that is employed by the CASU pipeline.
We have decided to enforce three basic criteria
which must be met for a candidate source
to be included in the catalogue:
\begin{enumerate}
\item the source must have been detected at S/N$>5$ in at least
one of the bands, i.e. it is required that at least one of
\emph{rErr}, \emph{iErr} or \emph{haErr} is smaller
than 0.2 mag;
\item the shape of the source must not be an obvious
cosmic ray or noise artefact, i.e. we require
either \emph{pStar} or \emph{pGalaxy} to be
greater than 20 per cent;
\item the source must not have been detected
in one of the strongly vignetted corners of the instrument, 
not have had any known bad pixels in the aperture,
and not have been on the edge of one of the CCDs.
\end{enumerate}

@@

Table~\ref{tbl:detections} details the breakdown of these
sources as a function of the bands in which they are captured.
159 million sources are detected in all three filters (73 per cent),
30 million are detected in two filters (14 per cent),
and the remaining 30 million are single-band detections.
Table~\ref{tbl:detections} also presents 
the fraction of `confirmed' objects,
which we define as those sources
which have been detected more than once
(recall that much of the survey area
is observed twice due to the field pair strategy).
We find that the single-band detections tend
to show much lower confirmation rates (32 per cent on average)
than double- and triple-band detections (89 per cent).
This suggests that a significant fraction of
these entries may be spurious detections.

@@

Not all the single-band detections are spurious, however.
We note that the confirmation rate for $i$-band detections
is markedly better than for $r$ and \ha,
which is likely to be explained by the fact that $i$ is least
affected by interstellar extinction, and so the survey can occasionally pick up
highly-reddened objects in $i$ which are otherwise lost in $r$ and \ha.
Moreover, objects which are intrinsically very red may also be picked
up in $i$ alone, while faint objects with very strong Balmer emission
may appear only in \ha.
Nevertheless, we recommend users not to rely on single-band objects
without inspecting the image data by eye,
or verifying that the object
was detected more than once (\emph{nObs}$>$1).

@@

@@

Towards Sh2-82,
the IPHAS and Spitzer/GLIMPSE catalogues
have 4,798 entries in common,
out of 10,739 and 11,321 entries in total, respectively.
Fig.~\ref{fig:spitzer-ccd} presents the Spitzer colour-colour diagram
of the region, showing the 1,356 objects that were detected 
in all four GLIMPSE bands.
We have indicated the 845 objects (62 per cent)
that have a counterpart in IPHAS (grey circles)
and the 511 objects that do not (red crosses).
